### 2017-01-24 Manuscript/SW_early_life_telomeres.Rmd
Home edits. Sent full draft to David.

###Abstract
1. Understanding the costs that individuals face throughout their lifetimes is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Investigating telomere dynamics may help us quantify individual variation in the costs of the social and ecological environment, and enhance our understanding of the dynamics of natural populations.

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*). We combine long-term life history and ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.

@@

The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015]. While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g. @Schultner2014; @Asghar2015]. There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016], which has important ramifications for our understanding of how telomeres reflect costs. However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.

@@

When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL[sample]; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL[individual]), using completely separate reactions run on separate plates. To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL[sample] and $\Delta$RTL[individual] using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL[inidividual] values were significantly different from $\Delta$RTL[sample] values, using Wilcoxon tests.

@@

For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL[sample] as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points. We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.

@@

There was positive correlation between telomere measurements taken at different time points (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL[individual] measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

@@

Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012]. However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period. Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance of individuals with shorter telomeres does not explain this pattern entirely. Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016]. The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015]. To our knowledge, ours is the first study to explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale. Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.

>>>>

###Abstract
1. Understanding individual-level variation in response to the social and ecological environment is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Investigating telomere dynamics may help us quantify individual variation in the costs of the social and ecological environment, and enhance our understanding of the dynamics of natural populations.

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*). We combine long-term life history and ecological data a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.

@@

The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015]. While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g. @Schultner2014; @Asghar2015]. There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs. However, increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.

@@

When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates. To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*inidividual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.

@@

For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL~*individual*~ as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points. We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.

@@

There was positive correlation between telomere measurements taken at different time points (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

@@

Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012]. However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period. Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance of individuals with shorter telomeres does not explain this pattern entirely. Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016]. The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015]. However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lenghtening, and that dismissing apparent telomere lengthening as solely measurment error is "too strong" without additional data [@Bateson2016a]. Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale. Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.

### 2017-01-24 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits. Sent full draft to David.

```{r Table 1,include = T,echo = F,warning = F}
RTL_Age_Dredge <- dredge(RTL_Age,
                         subset = !(LogAge & FAge) & 
                           !(LogAge & Agemonths) &
                           !(LogAge & Age2) &
                           !(Agemonths & FAge) & 
                           !(Agemonths & Age2) &
                           !(FAge & Age2))

@@

nams[nams == 'Agemonths'] <- 'Age (linear)'
nams[nams == 'LogAge'] <- 'Age (log)'
nams[nams == 'FAge'] <- 'Age (factor)'
nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
nams[nams == 'Agemonths:Cohort'] <- 'Age (continuous)*cohort'
nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
nams[nams == 'Age2'] <- 'Age (quadratic)'

@@

RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
                         subset = !(DeltaLogAge & DeltaAge) &
                           !(DeltaLogAge & DeltaAge2) &
                           !(DeltaAge & DeltaAge2))

@@

**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts. **A** RTL and age across all individuals. Points and  thin grey lines represent individual samples and birds, respectively. The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts. For visualisation purposes a selection of cohorts with large sample sizes across a wide age range was chosen, but all cohorts were included in models. **C** RTL and age among cohorts. Lines represent fitted values from a linear regression and log-transformed age, and colours correspond to **B**. **D** RTL in relation to and $\Delta$Log age (i.e. within indiviual variation in log age).

**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler. **A** Variation in RTL within individuals sampled at different time points. The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively. **B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual.

>>>>

```{r Table 1,include = T,echo = F,warning = F}
RTL_Age_Dredge <- dredge(RTL_Age,
                         subset = !(LogAge & FAge) & 
                           !(LogAge & Agemonths) &
                           !(LogAge & Age2) &
                           !(Agemonths & FAge) &
                           !(FAge & Age2) &
                           !(Age2 & !(Agemonths)))

@@

nams[nams == 'Agemonths'] <- 'Age (linear)'
nams[nams == 'MeanAge'] <- 'Age (mean)'
nams[nams == 'LogAge'] <- 'Age (log)'
nams[nams == 'FAge'] <- 'Age (factor)'
nams[nams == 'Cohort:FAge'] <- 'Age (factor)*cohort'
nams[nams == 'Agemonths:Cohort'] <- 'Age (linear)*cohort'
nams[nams == 'Cohort:LogAge'] <- 'Age (log)*cohort'
nams[nams == 'Age2:Cohort'] <- 'Age (quadratic)*cohort'
nams[nams == 'Age2'] <- 'Age (quadratic)'

@@

RTL_DeltaAge_Dredge <- dredge(RTL_DeltaAge,
                         subset = MeanAge &
                           !(DeltaLogAge & DeltaAge) &
                           !(DeltaLogAge & DeltaAge2) &
                           !(!(DeltaAge) & DeltaAge2))

@@

**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts. **A** RTL and age across all individuals. Points and  thin grey lines represent individual samples and birds, respectively. The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts. **C** RTL and age among cohorts. Lines represent fitted values from a linear regression and log-transformed age, and colours correspond to **B**. **D** RTL in relation to and $\Delta$Log age (i.e. within indiviual variation in log age).

**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler. **A** Variation in RTL within individuals sampled at different time points. The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively. **B** Scaled density plots of repeated RTL measurements among individual samples (grey), and among different samples taken from the same individual (yellow). Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases.
